---
title: "Class Reference"
author: "Isabelle Jensen"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
##trying to make changes
Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("C:/Users/izzoe/OneDrive/Documents/GitHub/Jensendata_journalism_2024_fall")

```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
install.packages("rmarkdown")
library(rmarkdown)
library(readr)
library(dplyr)
```


```{r}
maryland_winred_contributions <- read_rds("https://thescoop.org/files/maryland_winred.rds")
glimpse(maryland_winred_contributions)
head(maryland_winred_contributions)

#number of contributinos
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n()
  )

#number of contributions sorted high to low
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))

#date of contribution sorted by the number of contributions
maryland_winred_contributions |>
  group_by(memo_text, date) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))

#trump was convicted on May 30th so his contributions were much higher than usually on May 30th and May 31st

#funds sorted by total amount
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))

#median and mean sorted by amount of contribution
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount)
  ) |>
  arrange(desc(count_contribs))


#add min max abd sort from max amount
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
  arrange(desc(max_amount))

#largest donation
maryland_winred_contributions |>
  arrange(desc(amount))


#useful functions
#arrange() automatically groups numerical variables from least to greatest but using desc() function allows for greatest to least
#group_by() is helpful since it allows us to look data using the variable of interest and will be helpful for graphing
#sum() min() max() and others allow us to gain information about the data without having to ask about each input or group separately
```

