---
title: "Reverse Engineering Project"
author: "Isabelle Jensen"
date: "09-27-2024"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [Millions in out-of-state donations help fuel high-profile Maryland Democratic governor candidates](https://cnsmaryland.org/2022/03/03/millions-in-out-of-state-donations-help-fuel-high-profile-maryland-democratic-governor-candidates/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
Moore <- read.csv("data/Moore_ContributionsList.csv")
Perez <- read.csv("data/Perez_ContributionsList.csv")
King <- read.csv("data/King_ContributionsList.csv")
Franchot <- read.csv("data/Franchot_ContributionsList.csv")
Schulz <- read.csv("data/Schulz_ContributionsList.csv")
Gansler <- read.csv("data/Gansler_ContributionsList.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 
# This is so that we are able to organize donations by states, as detailed in the story.
# To do this, I am going to employ the following function.
?str_sub_all()
# all states are right before the zip code (-6 ch), and only have two letters (start at -8)
Moore  <-  Moore |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

Perez  <-  Perez |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

King  <-  King |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

Moore$Contributor.Address

unique(Moore$state)
unique(Perez$state)
unique(King$state)
glimpse(Moore)

#found some incorrect state names trying to fix
Moore <- Moore |> 
  mutate(state = str_extract(Contributor.Address, "\\b[A-Z]{2}(?= \\d{5}$)"))

Perez <- Perez |> 
  mutate(state = str_extract(Contributor.Address, "\\b[A-Z]{2}(?= \\d{5}$)"))

King <- King |> 
  mutate(state = str_extract(Contributor.Address, "\\b[A-Z]{2}(?= \\d{5}$)"))

Franchot <- Franchot |> 
  mutate(state = str_extract(Contributor.Address, "\\b[A-Z]{2}(?= \\d{5}$)"))

# Check the unique values again
unique(Moore$state)
unique(Perez$state)
unique(King$state)
unique(Franchot$state)

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Moore raised over $1 million in New York, and Perez gained 
over $226,000 from the state. Perez also received nearly $235,000 from California 
donors and raised over $100,000 in Florida and Virginia. Moore and King collected 
over $100,000 each from Illinois.” 


* **Analysis summary**:* For the Moore dataset, it is confirmed that he raised over 
$1 million in New York, and over $100,000 in Illinois. 
* For the Perez data, it is confirmed that he raised $226,208 in New York, almost
$235,000 in California, and over $100,000 in each Florida and Virignia.
* For the King data, it is confirmed that He raised over $100,000 from Illinois. 
* All of the data in this paragraph has been confirmed. After singling out the state
through str_sub function, all data has checked out as correctly represented in the story. 


```{r}
# Put code to reverse engineer sentence here
# Moore data: New York, Illinois
Moore |> 
  filter(state == "NY" | state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Moore raised exactly $1,002,915 from NY, and $131,629 from IL.

# Perez data: New York, California, Florida, Virginia
Perez |> 
  filter(state == "NY" | state == "CA" | state == "FL" | state == "VA") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Perez raised exactly $226,208.46 from NY, 
# $234,749 from California, $123,178 from VA, and $110,639 from FL.

# King data: Illinois
King |> 
  filter(state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# King raised $125,055.4 in Illinois

# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: Moore, a U.S. Army veteran and bestselling author, built a campaign war chest that included over $2.3 million in out-of-state contributions. Tom Perez, a former leader of the Democratic National Committee, raised nearly $1.7 million in donations from outside Maryland.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here


Moore_outofstate <- Moore |> 
  filter(state != "MD") |>            
  summarize(total_amount = sum(Contribution.Amount, na.rm = TRUE))
Moore_outofstate
#$2,386,593

Perez_outofstate <- Perez |> 
  filter(state != "MD") |>          
  summarize(total_amount = sum(Contribution.Amount, na.rm = TRUE))
Perez_outofstate
#$1,695,893


# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: John B. King, former secretary of Education under President Barack Obama, raised $2.2 million in campaign funding last year. Only 11%, just under $240,000, came from Maryland donors. King raised over $534,000 in New York alone.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here
Kingtotalfunding <- King |> 
  summarize(total_amount = sum(Contribution.Amount, na.rm = TRUE))
Kingtotalfunding
#$2,199,015

King |> 
  filter(state == "MD") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
#$238,092

238092/2199015
#0.1082721

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**:Moore, King and Perez stand in stark contrast to Peter Franchot, a former Maryland state delegate and current comptroller, who raised over 78% of his $1.66 million from inside Maryland.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

Franchottotalfunding <- Franchot |> 
  summarize(total_amount = sum(Contribution.Amount, na.rm = TRUE))
Franchottotalfunding
#total funding $1,663,516

Franchot |> 
  filter(state == "MD") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# Maryland funds $1,104,710

1104710/1663516	
#0.6640814
#potential problem with replication Im finding here that around 66% of Franchot's funds came from Maryland compared to what the article said which was 78%

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: “Less than a third of the money Perez raised, 31%, came from Maryland donors.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

Pereztotalfunding <- Perez |> 
  summarize(total_amount = sum(Contribution.Amount, na.rm = TRUE))
Pereztotalfunding
#total funding perez $2,458,688	


Perez |> 
  filter(state == "MD") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
#funding from maryland 736,344.4	

736344.4	/2458688	
#0.2994867 checks out less than a third from maryland

# Display results of code below this codeblock

```

-30-