---
title: "lab_03"
author: "Isabelle Jensen"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}

glimpse(md_contribs)
md_contribs <- md_contribs |>
  clean_names()

```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: There were misspellings or differences for Baltimore, Columbia, Potomac, Annapolis, Ellicott City, and Silver Spring. I fixed those and others to figure out which cities tryly had the most contributions. 

```{r}

md_contribs$contributor_city
sort(unique(md_contribs$contributor_city))

# BLATIMORE
#LUTHERVILLE TIMO and LUTHVLE TIMON and LUTHERVILLE TIMONI and LUTHERVILLE TIMONIUM and LUTHERVILLE-TIMONIUM
# MOUNT AIRY and MT. AIRY and MT.AIRY
#ANAPOLIS and ANNAPOLIS and ANNAPOLIS JUNCTION
# BEL AIR and BELAIR 
#BROOKEVILLE and BROOKVILLE
#COLUMBIA and COLOMBIA
# DAVIDSONVILLE and DAVISDSONVILLE
# ELLICOTT and ELLICOTT CITY
#GAITHERBURG and GAITHERSBURG
#MARRIOTSVILLE and MARRIOTTSVILLE
#N BETHESDA and NORTH BETHESDA
#OWINGS MILLS and OWINS MILLS
#POTAMIC and POTOMAC
#SILVER SPRING and SILVER SPRINGS
#SAINT MARYS CITY and ST MARYS CITY
#SAINT MICHAELS and ST MICHAELS
#UPPER MARLBORO and UPPR MARLBORO 

md_contribs <- md_contribs |>
  clean_names() |>  
  mutate(contributor_city = case_when(
    contributor_city == "BLATIMORE" ~ "BALTIMORE",
    contributor_city == "^LUTH" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "MT. AIRY" ~ "MOUNT AIRY",
    contributor_city == "MT.AIRY" ~ "MOUNT AIRY",
   contributor_city == "ANAPOLIS" ~ "ANNAPOLIS",
   contributor_city == "BELAIR" ~ "BEL AIR",
   contributor_city == "BROOKVILLE" ~ "BROOKEVILLE",
   contributor_city == "COLOMBIA" ~ "COLUMBIA",
    contributor_city == "DAVISDSONVILLE" ~ "DAVIDSONVILLE",
     contributor_city == "ELLICOTT" ~ "ELLICOTT CITY",
     contributor_city == "GAITHERBURG" ~ "GAITHERSBURG",
        contributor_city == "MARRIOTSVILLE" ~ "MARRIOTTSVILLE",
           contributor_city == "N BETHESDA" ~ "NORTH BETHESDA", 
     contributor_city == "OWINS MILLS" ~ "OWINGS MILLS", 
   contributor_city == "POTAMIC" ~ "POTOMAC", 
      contributor_city == "SILVER SPRINGS" ~ "SILVER SPRING", 
   contributor_city == "ST MARYS CITY" ~ "SAINT MARYS CITY", 
   contributor_city == "ST MICHAELS" ~ "SAINT MICHAELS", 
   contributor_city == "UPPR MARLBORO" ~ "UPPER MARLBORO", 
    TRUE ~ contributor_city
  ))

md_contribs |> 
  group_by(contributor_city) |> 
  summarise(
    count=n()
  ) |>
  arrange(desc(count))


```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: 20817 or parts of Montgomery county have the highest amount of contributions. It seems like a suburban area.  Helmer for Congress was the largest recipient of money from 20817. Two of the top five recipients are Maryland candidates.

```{r}

md_contribs <- md_contribs |>
  clean_names() |> 
  mutate(contributor_zip = str_sub(contributor_zip, start=1L, end=5L))

md_contribs$contributor_zip
  
md_contribs |> 
  group_by(contributor_zip) |> 
  summarise(
    total_amount = sum(contribution_receipt_amount),
    count = n()
  ) |> 
  arrange(desc(total_amount))



md_contribs |> 
  filter(contributor_zip == 20817) |> 
  group_by(committee_name) |> 
  summarise(
    total_amount = sum(contribution_receipt_amount),
    count = n()
  )|> 
  arrange(desc(total_amount))
```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: The top two donors are Mark Jacobsen, a banker from Bethesda, and Jeffery Drezner, an unemployed man from Potomax. In the top ten there is what is likely a duplicate because there is a Bill and a William Pugh from Bethesda that have donated the same amount. Its also worth noting that when I grouped the data also by occupation some people get listed twice because they have marked or had different jobs. In particular one of the top donors, Jeffery Drezner has marked unemployed and retired. 

```{r}
md_contribs |> 
  group_by(contributor_first_name, contributor_last_name, contributor_city ) |> 
  summarise(
    total_amount = sum(contribution_receipt_amount),
    count = n()
  )|> 
  arrange(desc(total_amount))


md_contribs |> 
  group_by(contributor_first_name, contributor_last_name, contributor_city, contributor_occupation ) |> 
  summarise(
    total_amount = sum(contribution_receipt_amount),
    count = n()
  )|> 
  arrange(desc(total_amount))

md_contribs |> 
  group_by(contributor_first_name, contributor_last_name, contributor_city ) |> 
  summarise(
    total_amount = sum(contribution_receipt_amount),
     contributor_occupation = first(contributor_occupation),
    count = n()
  )|> 
  arrange(desc(total_amount))
```
