---
title: "lab_07"
author: "derek willis"
date: "2024-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key *only* if you haven't already

```{r echo=FALSE}
census_api_key("c1593456bfe9d9813d469d3f50da2c0535a36217", overwrite=TRUE)
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2022 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2022 5-year ACS and save them to a dataframe

```{r}
acs5 <- load_variables(2022, "acs5", cache=TRUE)
View(acs5)






```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2022 5-year estimates? You will need to find the variable for median age first.

**A1** Baltimore county has the lowest median age at 36 years old. 

```{r}
age2022 <- get_acs(geography = "state", variables = "B01002_001", year = 2022)
#Estimate!!Median age --!!Total: Median Age by Sex

glimpse(age2022)
age2022

mdage <- get_acs(geography = "county",
              variables = c(medincome = "B01002_001"),
              state = "MD",
              year = 2022)
mdage

mdage |>
 group_by(NAME) |>
 arrange(estimate)

```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50.

**A2** Worcester County and Talbot County both have a median age above 50 with 50.6 and 51.4 respectively.

```{r}
mdage |>
 group_by(NAME) |>
  filter(estimate > 50)|>
 arrange(estimate)


```

## Ticket-Splitting

Ticket-splitting (voting for candidates from multiple parties) in today's political environment seems like a rarity, but it does happen, even in states like Maryland. For the remaining questions you'll be looking at cast vote records from Frederick County - Maryland's most competitive county - for the 2022 general election. You'll be trying to identify the zip codes where ticket-splitting is happening the most and to learn more about those places. To do that, you'll need to read in two CSV files, join them and then also bring in data from the Census API.

### Setup

Load the two CSV files in the `data` folder: one containing the ballot choices of Frederick voters in five races (governor, comptroller, attorney general and U.S. House) and the other containing information about Frederick's voting locations. You'll need to join them together.

```{r}

ballotchoice <- read_csv("data/frederick_cvr.csv")
precincts <- read_csv("data/frederick_precincts.csv")

ballotchoice
precincts


fredrick <- ballotchoice %>%
  inner_join(precincts, by = "precinct")

fredrick

```

**Q3** Let's answer some basic questions about ticket-splitting: how many voters voted for different parties in the governor and comptroller's race? How many voted for the same party in both races? What percentage of all voters split their tickets in those races? In how many cases did voters choose the same party in all five contests? What percentage of all voters does that represent?

**A3**Looking at all parties and write ins 11,871 voters voted for different parties in the governor and comptroller races. 95,221 voters voted for the same party for governor and comptroller. 11% of all voters split their ticket for comptroller and governor. 89,916 voters voted for the same party in all five races. About 84% of all voters voted for the same party in all 5 elections.  

```{r}

fredrick|>
  summarise(total_observations = n())
#total votes 107302

compgovsplit<-fredrick |>
 group_by(comptroller, governor_lt_governor) |>
  filter(governor_lt_governor != comptroller)|>
  summarize(
    count=n()
  )|>
 arrange(desc(count))
compgovsplit
sum(compgovsplit$count)
#11871


compgovsame<-fredrick |>
 group_by(comptroller, governor_lt_governor) |>
  filter(governor_lt_governor == comptroller)|>
  summarize(
    count=n()
  )|>
 arrange(desc(count))
compgovsame
sum(compgovsame$count)
#95221


11871/ 107302
#0.1106317

allsame<-fredrick |>
 group_by(comptroller, governor_lt_governor) |>
  filter(governor_lt_governor == comptroller, comptroller == attorney_general, attorney_general ==  u_s_senator, u_s_senator == representative_in_congress_6)|>
  summarize(
    count=n()
  )|>
 arrange(desc(count))
allsame
sum(allsame$count)
#89916


89916/107302
#0.8379713
```

**Q4** Now let's look for the zip codes where voters preferred Republicans in every race *except* the governor's race, where Dan Cox lost to Wes Moore by a wide margin. Write code to generate a dataframe of zip codes with the number of those voters for each of those zip codes, ordering the dataframe so that the zip code with the most ticket-splitters appears first.

**A4** 21702 has the most people who voted all republican except voting democrat for governor with 115 voters. There were 759 people total who voted this way.

```{r}
splitdemgov<-fredrick |>
 group_by(zipcode) |>
  filter(governor_lt_governor=="DEM", "REP" == comptroller, "REP" == attorney_general, "REP" ==  u_s_senator, "REP" == representative_in_congress_6)|>
  summarize(
    count=n()
  )|>
 arrange(desc(count))
splitdemgov

sum(splitdemgov$count)
#759


```

**Q5** Let's find out more about those zip codes from A4. Using tidycensus, get a dataframe of Maryland zip codes and their median ages, then join it to the dataframe you produced in A4. NOTE: you'll need to ensure that the two columns you use for the join are the same datatype. Summarize your findings; how would you describe the zip codes with more non-Cox Republicans compared to those with fewer? Where are those leading zip codes located?

**A5**

```{r}

zipage <- get_acs(geography = "zip code tabulation area",
                    variables = c(medincome = "B01002_001"),
                    year = 2022,
                    survey = "acs5")
colnames(zipage)

```






