---
title: "lab_04"
author: "Isabelle Jensen"
date: "2024-09-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
md_senate<- read_csv("data/md_senate_expenses.csv")

glimpse(md_senate)

```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.**

```{r}
new_md_senate<- read_csv("data/new-md-senate-expenses-csv.csv")

glimpse(new_md_senate)
```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.** Alsobrooks spent the most on Media production, Salary, List Acquistion, Online Advertising, Payroll Taxes, and Credit Card Payment. Hogan spent the most on Ad buy, Staffing, Travel Expense, Fundraising Services, and Strategy Consulting. They both spent a lot on ads. Hogan spend more on travel expenses. 

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.** For Alsobrooks I looked at Online Advertising spending. Alsobrooks started spending in May 2023 on online advertising. Spending was less than 10k every month until early 2024. In February 2024 there was a huge spike in online advertising spending. May and June 2024 saw the most spending at over 100k each month. I looked at a similar but less specific category for Hogan called Ad Buy. Hogan's spending on ads was very concentrated from March 2024 to June 2024. The campaign spent the most in May spending almost a million. This spike was short lived as spending was less than half that in June 2024. Both campaigns upped their ad spending in May probably because that was when the primary was. 

```{r}

new_md_senate |> 
  group_by(candidate, purpose_clean ) |> 
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(desc(total_amount))

new_md_senate |> 
  filter(purpose_clean == "Online Advertising") |> 
  filter(candidate == "Alsobrooks") |> 
  mutate(date = ymd(date),  
         month = floor_date(date, "month")) |>  
  group_by(month) |>  
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(month) 

new_md_senate |> 
  filter(purpose_clean == "AD BUY") |> 
  filter(candidate == "Hogan") |> 
  mutate(date = ymd(date),  
         month = floor_date(date, "month")) |>  
  group_by(month) |>  
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(month) 


```

**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.** I included salary, staffing, and payroll taxes. Alsobrooks spent a lot more on staffing, almost double. In these categories Alsobrooks paid 24 unique recipients. Hogan's spending all went to NA for categories of first and last name. Hogans spending in these categories all went to an organization called Purple Surfboard. Based on a short google, purple surfboard is something like a slogan for Hogan so he may have paid his employees through it. Regardless the data shows Hogan had 18 payments in this category to Alsobrooks 394. Something further is going on here whether I miss clustered the data or missed a category something about this doesn't seem right. I know Hogan had less money to spend but I wonder if some of the staffing for Alsobrooks includes things that Hogans does not. This could be due to the way they recorded it or the way I organized it. Or maybe Hogan paid lump sums of money to Purple Surfboard that they overtime distributed. More research is required to piece this out.  

```{r}
unique(new_md_senate$purpose_clean)



new_md_senate |> 
 filter(purpose_clean == "Salary" | purpose_clean == "STAFFING" | purpose_clean == "Payroll Taxes") |> 
   group_by(candidate) |> 
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(desc(total_amount))


new_md_senate |> 
 filter(purpose_clean == "Salary" | purpose_clean == "STAFFING" | purpose_clean == "Payroll Taxes") |> 
   group_by(candidate, last_name, first_name) |> 
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(candidate)

new_md_senate |> 
 filter(purpose_clean == "Salary" | purpose_clean == "STAFFING" | purpose_clean == "Payroll Taxes") |> 
   group_by(candidate, organization_name) |> 
  summarise(
    total_amount = sum(amount),
    count = n()
  ) |> 
  arrange(candidate)

```

**Q5** What's the most interesting story idea you've found in this data?

**A5** First of, Hogan spent less in general it seems. His biggest spending categories spent less than many of Alsobrooks. His funding is probably lower because it is such a democratic state so I wonder if more of his money comes from out of state than Alsobrooks. Second, the amount they pay their staff could be interesting and the way they pay them. Hogan using the Purple Surfboard organization rather than to individuals at least in the purposes I have seems odd. How normal is creating an organization to pay employees? Third, Hogan's travel expenses were way more than Alsobrooks. Why? Does Hogan tend to fly more or stay at nicer hotels or does he live farther away from more populated areas that require more campaigning. It looks like Hogan also spent more on events so maybe he holds more events than Alsobrooks does which means he has to travel more. This begs the question why does he hold more events and travel more than Alsobrooks? It might be different strategies of in person versus advertising but these strategies would be interesting to play out. 
